---
title: "Crime And Poverty Redesign Project"
date: 2025-09-25
categories: [Data Viz]
repo-url: https://github.com/jonathanwilsonami/stat-515
format: html
authors: 
  - "Jonathan M Wilson"
  - "Sean"
---

**Summary:** Data Visualization Redesign Project. 

::: {.badges}
**Tech Stack:** R, Python, Polars
:::

## Overview & Background 

Big Idea

3 Min Story
- Background 
- Original chart 
- Data used 
- Overview of data fetching process 

## Original Graphic

![The original chart[^refnote].](/images/old-chart-bubble-chart-crime.png){#fig-original fig-align="center" width="100%"}

[^refnote]: Gupta, A. (2022, February 23). 10 Good and Bad Examples of Data Visualization · Polymer. Polymer. https://www.polymersearch.com/blog/10-good-and-bad-examples-of-data-visualization

  The visualization shown above (see @fig-original) is the original graph that we chose for this redesign project. This visualization was taken from a website that identified multiple examples of data visualizations that do not follow best practices. Just at first glance, we can very quickly determine that there is a lot going on in this graph. In fact, there is too much going on to the point that it is hard for the audience to understand the intended story behind the graph. This graph attempts to showcase three separate variables in property crime, violent crime, and poverty, but did not do a great job in portraying the data in such a way that the audience can easily interpret. For example, the bubble radius for the poverty variable has no numeric reference, so even though they are comparable between states, any single radius has no meaning on its own. Additionally, there is no title of the graph so it is unclear what timeframe this data is coming from, the state labels are messy and hard to read, the units could be more explicit, and the color choice is not the best. Overall, there are several changes that can be made to this original graph to better illustrate the data and the message. 

  In our redesigns, the primary goal is to make the visualizations easier to understand than the original graph. This entails reducing the level of effort needed to interpret the visualizations. There are some elements of the original graph that had good intentions behind them though, that are worth incorporating in our redesigns. For example, the concept of tracking three variables in the same visualization can be insightful to examine their intertwined relationships. However, we can implement this concept more effectively by simplifying variable visuals and reducing overall visual clutter within the graph. Additionally, we can create multiple graphs side by side for comparison between these variables, with each graph focusing on only one or two variables. For another example, the concept of comparing variables between states is important in telling a story about the data. However, we can implement this concept more effectively by cleaning up the state labels and even utilizing a different element, such as a color gradient, to differentiate between the multiple poverty levels. Additionally, we can even move away from classic dot plots and explore plotting this data on a map, as most readers are familiar with a map of the US. These are just some of the various solutions that we implemented in our redesigns. 


```{r echo=FALSE, message=FALSE, warning=FALSE, error=FALSE}
# See renv.lock 
library(ggplot2)
library(dplyr)
# library(sf)
library(maps)
# library(stringr)
library(tidyr)
library(tidyverse) 
library(mapproj)
library(patchwork)
library(scales)
library(plotly)

##################################
# Data Sources 
##################################
state <- read.csv("data/crime_plus_poverty_2023.csv")
county <- read.csv("data/crime_poverty_by_county_2023.csv", check.names = FALSE)

##################################
# Data Cleaning 
##################################
# Droping rows where census data was not available for 2023 source county data 
if ("no_population" %in% names(county)) {
  county <- county %>% dplyr::filter(no_population != "true")
}

county <- county %>%
  mutate(
    poverty_rate = as.numeric(poverty_rate),
    property_crime_per_100k = as.numeric(property_crime_per_100k),
    violent_crime_per_100k = as.numeric(violent_crime_per_100k)
  )

##################################
# Useful functions & Vars
##################################
county_metric_cols <- c("poverty_rate",
            "property_crime_per_100k",
            "violent_crime_per_100k")

# Create geometries for mapping 
us_counties <- ggplot2::map_data("county")  # long/lat + region (state), subregion (county)
county_key  <- maps::county.fips %>% mutate(polyname = str_to_lower(polyname))

counties_poly <- us_counties %>%
  mutate(polyname = paste(region, subregion, sep = ",")) %>%
  left_join(county_key, by = "polyname") %>%
  rename(fips = fips) %>%
  filter(!is.na(fips))

# Function for creating maps
build_county_map <- function(data) {
  ggplot(data, aes(long, lat, group = group)) +
    geom_polygon(color = "gray85", linewidth = 0.1, na.rm = TRUE) +
    coord_quickmap() +
    theme_void(base_size = 12) +
    theme(
      legend.position = "bottom",
      plot.title = element_text(face = "bold", size = 16),
      plot.margin = margin(4, 8, 4, 8)
    )
}

# Helper to compute sensible upper limits (95th percentile) for crime and poverty rates 
p95 <- function(x) {
  x <- x[is.finite(x)]
  if (!length(x)) return(NA_real_)
  as.numeric(quantile(x, 0.95, na.rm = TRUE))
}

# Derive county FIPS given a data frame
# Uses: fips OR state+county codes OR state_abbr+county name
derive_fips <- function(df, county_key = NULL, keep_unmatched = FALSE) {
  stopifnot(is.data.frame(df))

  finish <- function(out) invisible(if (keep_unmatched) out else dplyr::filter(out, !is.na(.data$fips)))

  # county crosswalk (polyname = "state,county" lowercase)
  if (is.null(county_key)) {
    county_key <- maps::county.fips |>
      dplyr::mutate(polyname = stringr::str_to_lower(polyname))
  }

  # Already has FIPS
  if ("fips" %in% names(df)) {
    out <- dplyr::mutate(df, fips = stringr::str_pad(as.character(.data$fips), 5, "0"))
    return(finish(out))
  }

  # state + county numeric/string codes
  if (all(c("state", "county") %in% names(df))) {
    out <- dplyr::mutate(
      df,
      fips = paste0(
        stringr::str_pad(as.character(.data$state),  2, "0"),
        stringr::str_pad(as.character(.data$county), 3, "0")
      )
    )
    return(finish(out))
  }

  # state_abbr + county name (match via polyname)
  county_col <- c("county_name", "county")[c("county_name", "county") %in% names(df)][1]
  if (!is.na(county_col) && "state_abbr" %in% names(df)) {
    states_lookup <- tibble::tibble(
      state_abbr = c(state.abb, "DC"),
      state_name = c(state.name, "District of Columbia")
    ) |>
      dplyr::mutate(state_name_lower = stringr::str_to_lower(state_name))

    norm_county <- function(x) {
      x |>
        stringr::str_to_lower() |>
        stringr::str_replace_all("\\s+(county|parish|borough|census area|municipality|city)$", "") |> # Regex to extract county name
        stringr::str_replace_all("^st\\.?\\s", "saint ")
    }

    out <- df |>
      dplyr::left_join(states_lookup, by = "state_abbr") |>
      dplyr::mutate(polyname = paste(.data$state_name_lower, norm_county(.data[[county_col]]), sep = ",")) |>
      dplyr::left_join(county_key, by = "polyname")  # adds fips
    return(finish(out))
  }

  stop("derive_fips(): need one of: 'fips'; 'state'+'county'; or 'state_abbr'+county name.")
}

# Horizontal colorbar that sits nicely under the main title
nice_colorbar <- function() {
  guide_colorbar(
    title.position = "top",
    label.position = "bottom",
    barwidth = unit(90, "mm"),
    barheight = unit(5, "mm"),
    ticks = FALSE
  )
}

# Poverty ticks every 5 percentage points, capped to avoid clutter
poverty_breaks <- function(upper) {
  upper <- min(upper, 0.40)
  seq(0, upper, by = 0.05)
}

# Pretty numeric ticks up to the computed cap
pretty_breaks_to <- function(upper, n = 6) {
  scales::pretty_breaks(n)(c(0, upper))
}

# A cooler palette without the bright yellow; trim extremes
palette_args <- list(option = "mako", begin = 0.15, end = 0.90, direction = -1)
```

## Comparing Crime and Poverty by U.S. County 

```{r echo=FALSE, message=FALSE, warning=FALSE, error=FALSE}
#| fig-width: 18
#| fig-height: 24
#| fig-align: center
#| out-width: 100%
#| dpi: 200

df <- county

df <- derive_fips(df)

# Join metrics to polygons
map_df <- counties_poly %>%
  left_join(
    df %>% select(fips, all_of(county_metric_cols)),
    by = "fips"
  )

# Limits for each metric - 95th percentile as the upper scale limit
lim_pov  <- c(0, max(0.3, min(0.6, ceiling(p95(map_df$poverty_rate) * 10) / 10))) 
lim_prop <- c(0, p95(map_df$property_crime_per_100k)) 
lim_viol <- c(0, p95(map_df$violent_crime_per_100k)) 

poverty <- build_county_map(map_df) +
  geom_polygon(aes(fill = poverty_rate)) +
  do.call(scale_fill_viridis_c, c(
    palette_args,
    list(
      limits = lim_pov,
      breaks = poverty_breaks(lim_pov[2]),
      labels = scales::percent_format(accuracy = 1),
      oob = scales::squish, na.value = "gray95",
      name = "Poverty rate (% of residents)",
      guide = nice_colorbar()
    )
  )) +
  labs(title = "Poverty Rate by County") +
  theme(
    legend.position = "top",
    plot.title.position = "plot",
    legend.justification = "left",
    legend.box.just = "left",
    plot.title = element_text(size = 22, face = "bold")
  )

property <- build_county_map(map_df) +
  geom_polygon(aes(fill = property_crime_per_100k)) +
  do.call(scale_fill_viridis_c, c(
    palette_args,
    list(
      limits = lim_prop,
      breaks = pretty_breaks_to(lim_prop[2], n = 6),
      labels = scales::comma_format(accuracy = 1),
      oob = scales::squish, na.value = "gray95",
      name = "Property crime (per 100k residents)",
      guide = nice_colorbar()
    )
  )) +
  labs(title = "Property Crime per 100,000") +
  theme(
    legend.position = "top",
    plot.title.position = "plot",
    legend.justification = "left",
    legend.box.just = "left",
    plot.title = element_text(size = 22, face = "bold")
  )

violent <- build_county_map(map_df) +
  geom_polygon(aes(fill = violent_crime_per_100k)) +
  do.call(scale_fill_viridis_c, c(
    palette_args,
    list(
      limits = lim_viol,
      breaks = pretty_breaks_to(lim_viol[2], n = 6),
      labels = scales::comma_format(accuracy = 1),
      oob = scales::squish, na.value = "gray95",
      name = "Violent crime (per 100k residents)",
      guide = nice_colorbar()
    )
  )) +
  labs(title = "Violent Crime per 100,000") +
  theme(
    legend.position = "top",
    plot.title.position = "plot",
    legend.justification = "left",
    legend.box.just = "left",
    plot.title = element_text(size = 22, face = "bold")
  )

# Stack each plot vertically
stack_us <-
  (poverty / property / violent) &
  theme(
    plot.margin = margin(6, 10, 6, 10)
  )

stack_us +
  plot_annotation(
    title   = "Crime & Poverty by U.S. County",
    caption = "Note: Some county data are missing from the Census data\nSource: FBI Crime Data API and data.census.gov for 2023",
    theme = theme(
      plot.title.position = "plot",
      plot.title   = element_text(size = 28, face = "bold", hjust = 0), 
      plot.caption.position = "plot", 
      plot.caption = element_text(size = 16, hjust = 0, colour = "grey30",
                                  lineheight = 1.15, margin = margin(t = 10)),
      plot.margin  = margin(8, 12, 18, 12)
    )
  )
```

## Zooming in on South Carolina

```{r echo=FALSE, message=FALSE, warning=FALSE, error=FALSE}
#| fig-width: 14
#| fig-height: 20
#| fig-align: center
#| out-width: 100%
#| dpi: 200

df <- county

df <- derive_fips(df)

# Build South Carolina map data only
sc_poly <- counties_poly %>% filter(region == "south carolina")

sc_map <- sc_poly %>%
  left_join(df %>% select(fips, all_of(county_metric_cols)), by = "fips")

lim_pov  <- c(0, max(0.3, min(0.6, ceiling(p95(sc_map$poverty_rate) * 10) / 10)))
lim_prop <- c(0, {x <- p95(sc_map$property_crime_per_100k); if (!is.finite(x)) max(sc_map$property_crime_per_100k, na.rm = TRUE) else x})
lim_viol <- c(0, {x <- p95(sc_map$violent_crime_per_100k);  if (!is.finite(x)) max(sc_map$violent_crime_per_100k,  na.rm = TRUE) else x})

poverty_sc <- build_county_map(sc_map) +
  geom_polygon(aes(fill = poverty_rate)) +
  do.call(scale_fill_viridis_c, c(
    palette_args,
    list(
      limits = lim_pov,
      breaks = poverty_breaks(lim_pov[2]),
      labels = scales::percent_format(accuracy = 1),
      oob = scales::squish, na.value = "gray95",
      name = "Poverty rate (% of residents)",
      guide = nice_colorbar()
    )
  )) +
  labs(title = "Poverty Rate by County") +
  theme(
    legend.position = "top",
    plot.title.position = "plot",
    legend.justification = "left",
    legend.box.just = "left",
    plot.title = element_text(size = 22, face = "bold")
  )

property_sc <- build_county_map(sc_map) +
  geom_polygon(aes(fill = property_crime_per_100k)) +
  do.call(scale_fill_viridis_c, c(
    palette_args,
    list(
      limits = lim_prop,
      breaks = pretty_breaks_to(lim_prop[2], n = 6),
      labels = scales::comma_format(accuracy = 1),
      oob = scales::squish, na.value = "gray95",
      name = "Property crime (per 100k residents)",
      guide = nice_colorbar()
    )
  )) +
  labs(title = "Property Crime per 100,000") +
  theme(
    legend.position = "top",
    plot.title.position = "plot",
    legend.justification = "left",
    legend.box.just = "left",
    plot.title = element_text(size = 22, face = "bold")
  )

violent_sc <- build_county_map(sc_map) +
  geom_polygon(aes(fill = violent_crime_per_100k)) +
  do.call(scale_fill_viridis_c, c(
    palette_args,
    list(
      limits = lim_viol,
      breaks = pretty_breaks_to(lim_viol[2], n = 6),
      labels = scales::comma_format(accuracy = 1),
      oob = scales::squish, na.value = "gray95",
      name = "Violent crime (per 100k residents)",
      guide = nice_colorbar()
    )
  )) +
  labs(title = "Violent Crime per 100,000") +
  theme(
    legend.position = "top",
    plot.title.position = "plot",
    legend.justification = "left",
    legend.box.just = "left",
    plot.title = element_text(size = 22, face = "bold")
  )

# Stack the three maps
stack_sc <-
  (poverty_sc / property_sc / violent_sc) &
  theme(plot.margin = margin(6, 10, 6, 10))

stack_sc +
  plot_annotation(
    title   = "South Carolina by County",
    caption = "Note: Some county data are missing from the Census data\nSource: FBI Crime Data API and data.census.gov for 2023",
    theme = theme(
      plot.title.position = "plot",
      plot.title   = element_text(size = 28, face = "bold", hjust = 0),
      plot.caption.position = "plot",
      plot.caption = element_text(size = 16, hjust = 0, colour = "grey30",
                                  lineheight = 1.15, margin = margin(t = 10)),
      plot.margin  = margin(8, 12, 18, 12)
    )
  )

```

## Interactive Plot of South Carolina to Look at Individual Counties

```{r echo=FALSE, message=FALSE, warning=FALSE, error=FALSE}

# --- Load your data ---
df <- read.csv("data/crime_poverty_by_county_2023.csv", check.names = FALSE)

# optional flag -> logical
to_bool <- function(x) {
  if (is.logical(x)) return(ifelse(is.na(x), FALSE, x))
  if (is.numeric(x)) return(ifelse(is.na(x), FALSE, x != 0))
  if (is.character(x)) { y <- tolower(trimws(x)); return(y %in% c("true","t","1","yes","y")) }
  rep(FALSE, length(x))
}
if ("no_population" %in% names(df)) df <- df %>% filter(!to_bool(no_population))

df <- df %>%
  mutate(
    poverty_rate = as.numeric(poverty_rate),
    property_crime_per_100k = as.numeric(property_crime_per_100k),
    violent_crime_per_100k  = as.numeric(violent_crime_per_100k)
  )

# --- County polygons and FIPS via maps::county.fips ---
us_counties <- ggplot2::map_data("county")
county_key  <- maps::county.fips %>% mutate(polyname = str_to_lower(polyname))

counties_poly <- us_counties %>%
  mutate(polyname = paste(region, subregion, sep = ",")) %>%
  left_join(county_key, by = "polyname") %>%
  rename(fips = fips) %>%
  filter(!is.na(fips))

# If your CSV doesn’t have fips, try to derive from names (state_abbr + county name)
if (!"fips" %in% names(df)) {
  county_col <- if ("county_name" %in% names(df)) "county_name"
                else if ("county" %in% names(df)) "county" else NA_character_
  if (!is.na(county_col) && "state_abbr" %in% names(df)) {
    states_lookup <- tibble(
      state_abbr = c(state.abb, "DC"),
      state_name = c(state.name, "District of Columbia")
    ) %>% mutate(state_name_lower = str_to_lower(state_name))
    df <- df %>%
      left_join(states_lookup, by = "state_abbr") %>%
      mutate(
        state_name_norm  = state_name_lower,
        county_name_norm = str_to_lower(.data[[county_col]]) %>%
          str_replace_all("\\s+(county|parish|borough|census area|municipality|city)$", "") %>%
          str_replace_all("^st\\.?\\s", "saint "),
        polyname = paste(state_name_norm, county_name_norm, sep = ",")
      ) %>%
      left_join(county_key, by = "polyname") %>%
      rename(fips = fips)
  }
}
df <- df %>% filter(!is.na(fips))

# --- Build SC-only map data ---
sc <- counties_poly %>% filter(region == "south carolina") %>%
  left_join(df %>% select(fips, all_of(county_metric_cols)), by = "fips") %>%
  mutate(
    hover_text = paste0(
      "County: ", str_to_title(subregion), "<br>",
      "Poverty rate: ", percent(poverty_rate, accuracy = 0.1), "<br>",
      "Property /100k: ", comma(property_crime_per_100k, accuracy = 1), "<br>",
      "Violent /100k: ",  comma(violent_crime_per_100k,  accuracy = 1)
    )
  )

# Pick which metric to color by
fill_var <- "violent_crime_per_100k"  # or "poverty_rate" or "property_crime_per_100k"

# sensible upper limit (95th pct) + labels
p95 <- function(x) { x <- x[is.finite(x)]; if (!length(x)) return(NA_real_); as.numeric(quantile(x, 0.95, na.rm = TRUE)) }
if (fill_var == "poverty_rate") {
  lim_hi <- max(0.3, min(0.6, ceiling(p95(sc$poverty_rate) * 10) / 10))
  label_fun <- scales::percent_format(accuracy = 1)
  legend_name <- "Poverty"
  title_txt <- "South Carolina — Poverty Rate"
} else if (fill_var == "property_crime_per_100k") {
  lim_hi <- p95(sc$property_crime_per_100k); if (!is.finite(lim_hi)) lim_hi <- max(sc$property_crime_per_100k, na.rm = TRUE)
  label_fun <- scales::comma_format(accuracy = 1)
  legend_name <- "Per 100k"
  title_txt <- "South Carolina — Property Crime per 100k"
} else {
  lim_hi <- p95(sc$violent_crime_per_100k); if (!is.finite(lim_hi)) lim_hi <- max(sc$violent_crime_per_100k, na.rm = TRUE)
  label_fun <- scales::comma_format(accuracy = 1)
  legend_name <- "Per 100k"
  title_txt <- "South Carolina — Violent Crime per 100k"
}

# --- Plot (ggplot -> plotly) ---
p_sc <- ggplot(sc, aes(long, lat, group = group)) +
  geom_polygon(aes(fill = .data[[fill_var]], text = hover_text), color = "gray85", linewidth = 0.1, na.rm = TRUE) +
  coord_quickmap() +
  scale_fill_viridis_c(
    labels = label_fun,
    limits = c(0, lim_hi),
    oob = scales::squish,
    na.value = "gray95",
    direction = -1,
    name = legend_name
  ) +
  labs(title = title_txt) +
  theme_void(base_size = 12) +
  theme(legend.position = "bottom", plot.title = element_text(face = "bold", size = 18))

plotly::ggplotly(p_sc, tooltip = "text") %>%
  plotly::layout(hovermode = "closest")

```

